---
title: "NF-OSI Data Access Survey -- 2023 Preliminary Results"
description: |
  Preliminary survey findings from the NF community on the experience of data discovery, sharing, and reuse through the NF Data Portal.
author:
  - name: Anh Nguyet Vu
    url: https://github.com/anngvu
    affiliation: Sage Bionetworks
    affiliation_url: https://sagebionetworks.org/
    orcid_id: 0000-0003-1488-6730
  - name: Jineta Banerjee
    url: https://github.com/jaybee84
    affiliation: Sage Bionetworks
    affiliation_url: https://sagebionetworks.org/
    orcid_id: 0000-0002-1775-3645
  - name: Christina Conrad
    url: https://github.com/cconrad8
    affiliation: Sage Bionetworks
    affiliation_url: https://sagebionetworks.org/
    orcid_id: 0000-0001-8688-2523
  - name: Robert Allaway
    url: https://github.com/allaway
    affiliation: Sage Bionetworks
    affiliation_url: https://sagebionetworks.org/
    orcid_id: 0000-0003-3573-3565
date: "`r Sys.Date()`"
output: distill::distill_article
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(dplyr)
library(tidyr)
library(ggplot2)
library(gt)
library(googlesheets4)
library(tidygraph)
library(ggraph)

source("R/circlepack.R")
```


```{r, include=FALSE}
sheet_id <- "1mZ6_XzC1d82cCl8uBqBjSDvIMo7-QN4Q2_FMIifbmkA"
raw <- googlesheets4::read_sheet(ss = sheet_id)

```

```{r, echo=F}
# users
emails <- raw %>%
  pull(`Email Address`) %>%
  unique()

deid <- setNames(1:length(emails), emails)
survey <- raw %>%
  mutate(id = deid[`Email Address`]) %>%
  select(-c(`Email Address`))

# use only most recent response where some users submitted multiple responses
survey <- survey %>%
   group_by(id)  %>%
   top_n(n = 1, Timestamp)
```
## Summary:
## Introduction

Since 2015, the NF-OSI data platform has continued to grow and is now the entrypoint to over two research studies for the Neurofibromatosis (NF) research community.
Eight years later, with the support of the Neurofibromatosis Therapeutic Acceleration Program (NTAP), we surveyed NF community members on the experience of **data findability, sharing, and reuse** through the NF Data Portal. The survey aimed to evaluate usability or barriers thereof in these three key areas to understand where the most impactful improvements could be made.

## Results and Analysis

### Respondent profiles
```{r fig-respondent-type, fig.cap="Demographic representation of survey respondents", layout="l-body-outset", fig.width=9, fig.height=7, echo=F}

# Format labels
survey$`What is your professional title?` <- gsub("Bioinformatician/computational biologist", "Bioinformatician\n/ comp. biologist", survey$`What is your professional title?`)
survey$`What is your professional title?` <- gsub("Bench researcher/staff scientist", "Bench researcher\n/ staff scientist", survey$`What is your professional title?`)
type_representation(`What is your professional title?`, survey)

```

Our sample of respondents (*n=`r nrow(survey)`*) represented different types of users.
Figure \@ref(fig:fig-respondent-type) summarizes the relative abundance of six demographic types.
Because respondents could indicate more than one demographic type, a statistic of 54% here should be interpreted to mean that 54% of the responses represented the "Principal Investigator" profile.
Immediately evident is the that fact that the respondents contain a large representation of Principal Investigator (PIs), more than for any other type of user. 
This is very encouraging and suggests that the NF Data Portal and the survey is reaching community constituents who have significant say in how research should be done, e.g. how to both obtain and share data for a study.

Conversely, a relatively low percentage of bioinformaticians/computational biologists were noted among respondents. 
This proportion for bioinformaticians/computational biologists is also about the same as a [previous RFC survey](https://nf-osi.github.io/research/rfc-brief.html#respondent-profiles).
A data portal should arguably see *over-representation* of bioinformaticians in its userbase, suggesting that the Portal either needs to have more bioinformatician users or increase survey engagement for this group.
<!-- read: computational people may fill out surveys at a lower rate -->
The second issue is that the survey question did not adequately admit additional demographics.
One other type of user that should perhaps have been captured is the "rare disease patient or patient advocate".
However, we did see that a small minority of respondents belonging to this demographic on further investigation of respondent profiles. 
It is highly encouraging to see that patients or patient advocates are also aware and are engaging with the NF Data Portal.
We intend to capture the thoughts of this demographic more systematically in our next versions of this survey.

```{r reshape, echo=FALSE}

# long format (tidy) data needed for subsequent figures

tidy_survey_quant <- survey %>%
  select(starts_with("How would you generally rate your experience")) %>%
  tidyr::pivot_longer(cols = !id, names_to = "area", values_to = "rating")

area_regex <- "(ability to \\w+)"
issue_regex <-"(?<=\\[).+?(?=\\])"

tidy_survey_ordinal <- survey %>%
  select(contains("sort")) %>%
  tidyr::pivot_longer(cols = !id, names_to = "area_issue", values_to = "response") %>%
  mutate(area = regmatches(area_issue, regexpr(area_regex, area_issue, perl = T)),
         issue = regmatches(area_issue, regexpr(issue_regex, area_issue, perl = T)))


grouped_bar <- function(data) {

  ggplot(data, aes(x = issue, y = n, fill = response)) +
       geom_bar(stat="identity") +
  scale_fill_manual(values = c("#E64B35FF",  "orange", "#4DBBD5FF", "gray")) +
  theme_minimal() +
  xlab("") +
  ylab("Number of responses") +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1),
        plot.margin = margin(3, 1, 0.4, 1, "cm")) # need generous left margin for long text
}

```

### Data findability

Data discovery is an important antecedent to data reuse.
The first survey component asked about issues community users might encounter when trying to **find** data through the NF Data Portal.

Of several issues that might hamper findability, Figure \@ref(fig:fig-findability) indicates that a confusing user interface had the most severe effect, though poor search functionality was also detrimental.
Half or nearly half of users found it "hard" or "impossible" to find data because of these two issues.

```{r fig-findability, fig.cap="Caption", layout="l-body-outset", fig.width=9, fig.height=7}

findability <- tidy_survey_ordinal %>%
  filter(area == "ability to find") %>%
  mutate(response = factor(response, levels = c("Made it impossible to find any data", "Made it hard to find data", "This was not an issue", "Not applicable"))) %>%
  group_by(issue) %>%
  count(response)

grouped_bar(findability)

```


### Data reuse

The second part of the survey asked community users to reveal what might have negatively impacted their ability to **reuse** data.
Within this area, there were two quite different potential issues: the technical issue of transfer capacity versus the more business-operational issue of dealing with an embargo period.
## Major focus area of improvement identified:
 1. Data transfer rate and capacity from the Portal UI

```{r fig-reuse, fig.cap="Caption", layout="l-body-outset", fig.width=9, fig.height=7}

download <- tidy_survey_ordinal %>%
  filter(area == "ability to download") %>%
  mutate(response = factor(response, levels = c("Made it impossible to download any data", "Made it hard to download data", "This was not an issue", "Not applicable"))) %>%
  group_by(issue) %>%
  count(response)

grouped_bar(download)

```

### Data sharing

Data sharing through the NF platform includes not only uploading data but also providing metadata (annotations) for the data, thus this survey section asked about the experience of uploading and using our annotation tools.

Figure \@ref(fig:fig-sharing-upload) shows that the issue which made upload "hard" or "impossible" for the largest share of users was "confusing or insufficient instructions".
Even though other issues affected a smaller share of users, they still represent major blockers for those users who said these made upload "impossible".

```{r fig-sharing-upload, fig.cap="Caption", layout="l-body-outset", fig.width=9, fig.height=7}
upload <- tidy_survey_ordinal %>%
  filter(area %in% c("ability to upload")) %>%
   mutate(response = factor(response, levels = c("Made it impossible to upload", "Made it hard to upload data", "This was not an issue", "Not applicable"))) %>%
   group_by(issue) %>%
  count(response)

grouped_bar(upload)
```

As with data upload, the majority and largest proportion of users thought "confusing or insufficient instructions" was a real issue (Figure \@ref(fig:fig-sharing-annotation)).
## Major focus area of improvement identified:
 1. Data annotation is a major hurdle for users
 2. Lack of instructions is a major barrier for smooth data annotation


```{r fig-sharing-annotation, fig.cap="Caption", layout="l-body-outset", fig.width=9, fig.height=7}
annotation <- tidy_survey_ordinal %>%
  filter(area == "ability to annotate") %>%
  mutate(response = factor(response, levels = c("Made it impossible to annotate", "Made it hard to annotate", "This was not an issue", "Not applicable"))) %>%
   group_by(issue) %>%
  count(response)

grouped_bar(annotation)
```


### User sentiment

Our results indicated that there are some users being left behind in all key areas.

Users were also asked to provide a rating to summarize their overall experience in each area.
When comparing the response distribution for these different areas all together, we see somewhat interestingly that the responses tend to be bi-modal (Figure \@ref(fig:fig-all-distributions)).
Users who respond appear to be really pretty dissatisfied or pretty satisfied.
While there is no area that is truly terrible, data sharing does have the lowest overall rating, with metadata sharing through the Data Curator App does having the lowest experience rating of 3 (Table \@ref(tab:table-ratings)).

To improve the data sharing experience, the analysis has already suggested having clearer instructions and making sure they are seen by users.
However, user comments also provide the insight that instructions become an issue because the current order/decoupling of events for data sharing is unintuitive -- data annotation should be done with data upload, not after.
Thus it may not be really that more instructions are needed but that the process could be simpler and more intuitive.
This would require both a technological and operational improvement to better combine data upload and annotation.

Given that data sharing should be prioritized first, the next prioritization chooses between data findability versus data reuse.
The survey results suggests that data findability affected more users.
Both search functionality and the user interface are issues to target for data findability.


```{r fig-all-distributions, fig.label="", fig.width=6.5, fig.height=10, echo=F}

h <- tidy_survey_quant %>%
  mutate(area = gsub("\\(.*", "", area))

ggplot(h, aes(x = rating)) +
  geom_histogram(binwidth=1) +
  facet_wrap(area ~ ., ncol = 1) +
  theme_light()
```


```{r table-ratings, fig.label="", echo=F}

medians <- tidy_survey_quant %>%
  group_by(area) %>%
  summarize(median = round(median(rating), 2))

 gt(medians)

```



## Other discussion

Outside of the main survey, additional comments left by respondents indicated that improving data accessibility to patients is also needed for the NF Data Portal.


## Continuing research

These preliminary results helped to establish a baseline for our community's experience and to recommend the right allocation of resources for improvement.
Nevertheless, the modest sample sizes could have possibly skewed responses for certain areas and types of users.
The [portal survey](https://docs.google.com/forms/d/e/1FAIpQLSdSgkq66IoLHbvXNmMEjEg4nMELwM-_CaJK3rFkU9pn84gYuA/viewform) remains open to collect more data.
To make it more inclusive we will revise the respondent profiling question to include other demographics options. 
In addition to getting more representative sample sizes, the continuation of the survey will allow trend analysis to see how experiences change over time. 
The results will be integrated into plans for continuous improvement in data discovery, sharing, and reuse as the NF data platform (hopefully) continues to expand and receive investment.

The survey is also somewhat restricted in the insights that could be obtained because of greater focus on quantitative data and specific issues.
More open-ended and qualitative research could extend the research to uncovered issues as well as explore ideas on how to build a better portal.
A number of respondents indicated that they were willing to participate in interviews as part of this more qualitative exploration, which might be done through partnership with the Sage design team.
